---
title: >
  The `EMMA` User's Guide
author:
- name: Najla Abassi
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  email: abassina@uni-mainz.de
- name: Annekathrin Nedwed
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  email: anneludt@uni-mainz.de
- name: Federico Marini
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  - Research Center for Immunotherapy (FZI), Mainz
  email: marinif@uni-mainz.de
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('EMMA')`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    code_folding: show
    code_download: yes
vignette: >
  %\VignetteIndexEntry{Introduction_to_EMMA}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{EMMA}
  %\VignetteKeywords{Pathways, GeneSetEnrichment, Annotation, ReproducibleResearch, Transcriptomics}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: EMMA_bibliography.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment  = "#>",
  error    = FALSE,
  warning  = FALSE,
  eval     = FALSE,
  message  = FALSE
)
```

# Introduction {#introduction}

Functional Enrichment Analysis (FEA) is a key downstream step in omics workflows,
commonly applied after differential expression analysis to support biological
interpretation and generate pathway-level hypotheses. A wide range of tools and
methods are available, leading to substantial heterogeneity in analytical choices
and reported results.

Despite its widespread use, FEA is often insufficiently documented. Critical
parameters such as background gene sets or multiple testing correction methods
are frequently missing or inconsistently reported in scientific papers,
limiting reproducibility and interpretability. Currently, no standardized method
exists to ensure transparent and reproducible documentation of FEA workflows.

To address this gap, we introduce `r BiocStyle::Biocpkg("EMMA")`, a framework
that automatically captures key analytical parameters and provenance information
during the execution of FEA methods.

This vignette demonstrates how `EMMA` integrates with existing tools
(e.g. `clusterProfiler`, `topGO`, `Enrichr`, `gprofiler2`) to execute enrichment
analyses while systematically capturing analysis parameters and provenance
information during runtime, and returning enrichment results in their native
format alongside structured and reusable metadata.


# Getting started {#gettingstarted}

To install the development version of this package, start R and enter:

```{r install, eval = FALSE}
library("remotes")
remotes::install_github("imbeimainz/EMMA",
                        dependencies = TRUE,
                        build_vignettes = TRUE)
```

Once installed, the package can be loaded and attached to the current workspace
as follows:

```{r setup}
library("EMMA")
```

# `EMMA` on the `macrophage` dataset

In the remainder of this vignette, we will illustrate the main features of `r BiocStyle::Biocpkg("EMMA")` on a publicly available dataset from
Alasoo, et al. "Shared genetic effects on chromatin and gene expression indicate
a role for enhancer priming in immune response", published in Nature Genetics,
January 2018 
[@Alasoo2018] [doi:10.1038/s41588-018-0046-7](https://doi.org/10.1038/s41588-018-0046-7).

The data is made available via the `r BiocStyle::Biocpkg("macrophage")`
Bioconductor package, which contains the files output from the Salmon
quantification (version 0.12.0, with GENCODE v29 reference), as well as the
values summarized at the gene level, which we will use to exemplify.

In the `macrophage` experimental setting, the samples are available from 6
different donors, in 4 different conditions (naive, treated with Interferon
gamma, with SL1344, or with a combination of Interferon gamma and SL1344).

Let's start by loading all the necessary packages:

```{r loadlibraries}
library("EMMA")

library("macrophage")
library("DESeq2")
library("org.Hs.eg.db")
library("clusterProfiler")
```

We will show an example of how  `r BiocStyle::Biocpkg("EMMA")` fits into a
regular bulk RNA-seq data analysis workflow.

# Get a list of Differentially Expressed Genes

For this, we will load the `macrophage` data and perform Differential Expression
Analysis with `r BiocStyle::Biocpkg("DESeq2")`

```{r dds}
# load data
data(gse, "macrophage")
# set up design
dds_macrophage <- DESeqDataSet(gse, design = ~ line + condition)
# preprocess
rownames(dds_macrophage) <- substr(rownames(dds_macrophage), 1, 15)
keep <- rowSums(counts(dds_macrophage) >= 10) >= 6

dds_macrophage <- dds_macrophage[keep, ]
dds_macrophage
```

We then run the main `DESeq()` function and check the `resultsNames` being
generated:

```{r run_deseq}
# run DESeq
dds_macrophage <- DESeq(dds_macrophage)
# contrasts
resultsNames(dds_macrophage)
```

Let's extract the DE results for each contrast. We will do three contrasts:

```{r get_de_res}
# get de res for 1st contrast
IFNg_vs_naive <- results(dds_macrophage,
                         contrast = c("condition", "IFNg", "naive"),
                         lfcThreshold = 1, alpha = 0.05)
IFNg_vs_naive <- lfcShrink(dds_macrophage, coef = "condition_IFNg_vs_naive",
                           res = IFNg_vs_naive,
                           type = "apeglm")
IFNg_vs_naive$SYMBOL <- rowData(dds_macrophage)$SYMBOL
IFNg_vs_naive

# get de res for 2nd contrast
Salm_vs_naive <- results(dds_macrophage,
                         contrast = c("condition", "SL1344", "naive"),
                         lfcThreshold = 1, alpha = 0.05)
Salm_vs_naive <- lfcShrink(dds_macrophage, coef = "condition_SL1344_vs_naive",
                           res = Salm_vs_naive,
                           type = "apeglm")
Salm_vs_naive$SYMBOL <- rowData(dds_macrophage)$SYMBOL
Salm_vs_naive

# get de res for 3rd contrast
Both_vs_naive <- results(dds_macrophage,
                         contrast = c("condition", "IFNg_SL1344", "naive"),
                         lfcThreshold = 1, alpha = 0.05)
Both_vs_naive <- lfcShrink(dds_macrophage,
                           coef = "condition_IFNg_SL1344_vs_naive",
                           res = Both_vs_naive,
                           type = "apeglm")
Both_vs_naive$SYMBOL <- rowData(dds_macrophage)$SYMBOL
Both_vs_naive
```

Now let's sort the results based on the adjusted p-value:

```{r}
# have a list of de results per contrast
contrast_list <- list(IFNg_vs_naive = IFNg_vs_naive,
                      Salm_vs_naive = Salm_vs_naive,
                      Both_vs_naive = Both_vs_naive)

de_res_list <- list()
# sort
for (i in names(contrast_list)) {
  de_res <- as.data.frame(contrast_list[[i]])
  de_res <- de_res[order(de_res$padj), ]
  #de_res <- de_res[!(is.na(de_res$padj)) & de_res$padj <= 0.05, ]
  
  de_res_list[[i]] <- de_res
}

# set universe
universe <- rownames(dds_macrophage)
```

# Perform Functional Enrichment Analysis (FEA)

Now that we have a list of DE genes for each contrast, we can perform Functional
Enrichment Analysis. For the first contrast `IFNg_vs_naive`, we will use the
function `enrichGO()` from `r BiocStyle::Biocpkg("clusterProfiler")`

`EMMA_run()` accepts a function call (e.g. `enrichGO(...)`) and executes it as
it is, while capturing the associated parameters and provenance information:

```{r EMMA_run_1}
# get de res from the list
de_res_IFNg_vs_naive <- de_res_list$IFNg_vs_naive

# perform FEA, but with EMMA!
fea_res <- enrichGO(gene = rownames(de_res_IFNg_vs_naive),
                    keyType = "ENSEMBL",
                    OrgDb = org.Hs.eg.db,
                    ont = "BP",
                    pAdjustMethod = "BH",
                    pvalueCutoff = 0.05,
                    qvalueCutoff = 0.1,
                    universe = universe) |> 
  EMMA_run() # simply pipe your call to EMMA_run()
```

... or you can simply wrap `EMMA_run()` around your call:

```{r EMMA_run_2}
# you can also pass the function name and its namespace
# e.g. `clusterProfiler::enrichGO(...)`
fea_res <- EMMA_run(enrichGO(gene = rownames(de_res_IFNg_vs_naive),
                    keyType = "ENSEMBL",
                    OrgDb = org.Hs.eg.db,
                    ont = "BP",
                    pAdjustMethod = "BH",
                    pvalueCutoff = 0.05,
                    qvalueCutoff = 0.1,
                    readable = TRUE))

# check res
fea_res
```

As you can see, `EMMA` returns the FEA results in their **native/standard** format.
`EMMA` also warns you about good practices when performing FEA, like in this
example, we didn't define a list of background genes, so we get warned about that.

To get a quick summary of what `EMMA` captured while we ran the analysis, we use
`EMMA_show()`:

```{r EMMA_show}
EMMA_show(fea_res)
```

`EMMA` stores the captured metadata within the attributes of the results object.
That's why it is always a good practice to save the original results, and not
only the subsets of interest.
To see the full recorded information, we use `getEMMARecord()`:

```{r getEMMARecord}
emma_record <- getEMMARecord(fea_res)

emma_record
```

`EMMA` aims to capture the following elements:

```
├── EMMA_record
│   ├── method
│   │   ├── call
│   │   └── function_name
│   │   └── package_name
│   │   └── package_version
│   └── input
│   │   └── arguments
│   └── annotation
│   │   ├── organism
│   │   └── gene_set_db
│   │   └── gene_set_db_version
│   └── timestamp
│   └── session_info
│   └── user_metadata # the user can pass any extra metadata freely in this field
│   ├── emma_version
```

As we can see a lot of information was recorded. With `EMMA_run()`, we can decide
whether we want to save the value of arguments used in our call or not. For this,
we can use the argument `args_form`:

```{r argument_form}
fea_res_no_param <- enrichGO(gene = rownames(de_res_IFNg_vs_naive),
                             universe = universe,
                             keyType = "ENSEMBL",
                             OrgDb = org.Hs.eg.db,
                             ont = "BP",
                             pAdjustMethod = "BH",
                             pvalueCutoff = 0.05,
                             qvalueCutoff = 0.1,
                             readable = TRUE) |> 
  EMMA_run(args_form = "unevaluated") # if we don't want the values stored
                                      # else set to evaluated (default)

# check
getEMMARecord(fea_res_no_param)
```

We can also choose whether to save the session info with the record using the 
argument `session`, which defaults to true.

`EMMA_freeze()` records the R environment during analysis runtime and generates
a lockfile that can be used with tools such as `renv`, to facilitate
reproducible reconstruction of the analysis environment.

```{r EMMA_freeze}
EMMA_freeze()
```

`EMMA_explain()` generates a human-readable description of the FEA, similar
to a Materials and Methods section of a paper, by summarizing the executed call,
the parameters, software context, and reference databases used.

```{r EMMA_explain}
EMMA_explain(fea_res)
```

# Session info {.unnumbered .smaller}

```{r sessioinfo}
sessionInfo()
```

# References {.unnumbered}
